---
title: "Big 3 Results"
author: "Ben Gonzalez"
date: '2023-10-19'
output: html_document
---

# Big Three Results {-}

<center><h1>
Statistical Significance - Confidence Intervals - Effect Size
</h1>
</center>




```{r echo=FALSE, include=TRUE}

knitr::opts_chunk$set(echo = TRUE)
```

## Sample Standard Deviation Formula {-}

$$\large
S\;=\;\sqrt{\frac{\sum(X\;-\;{\overline{X}})^2}{N\;-1}}
$$

```{r}

N <- length(mtcars$hp)
deviations <- mtcars$hp-mean(mtcars$hp)

s <- deviations^2

m_plus<- sum(s)/(N-1)

sd_plus <- sqrt(m_plus)
print(sd_plus)

sqrt((sum((mtcars$hp-mean(mtcars$hp))^2))/(length(mtcars$hp)-1))

sd(mtcars$hp)
```




## Standard Error Formula {-}

$$\large
SE\;\;=\frac{\sigma}{\sqrt{n}}
$$
$$
SE\;=Standard\;Error
\\
\sigma\;=sample\;standard\;deviation
\\
n\;=number\;of\;samples
$$

## How to calculate the standard error of a sample {-}

```{r}
#standard deviation/squareroot(n)

# length(mtcars$hp)
# nrow(mtcars)

###Shortcut to calculate the standard error of a sample
###The length function is utilized to find the number of observations in a data set
sd(mtcars$hp)/sqrt(length(mtcars$hp))


####Another shortcut to calculate the standard error of a sample
###The nrow function is utilized to find the number of observations in a data set
sd(mtcars$hp)/sqrt(nrow(mtcars))

####Long way to calculate the standard error of a sample
print(sqrt(sum((mtcars$hp - mean(mtcars$hp)) ^ 2/(length(mtcars$hp) - 1)))
      /sqrt(length(mtcars$hp)))

```


## Computing Confidence Intervals {-}


<br>
<br>

## Confidence Interval Formula {-}

The confidence interval (C.I.) according to (Hatcher, 2013) gives us a range of values for the population parameter being estimated.

Computed for:

+ mean
+ difference between means
+ correlation coefficients
+ etc.

## Lower Bound of the C.I. for a Sample Mean {-}

$$\large
CI\;=\;\overline{X}\;\pm\;(SE_{m})(t_{crit})
$$

$$
CI\;=Confidence\;Interval
\\
\overline{X}\;=\;observed\;sample\;mean
\\
SE_{m}\;=standard\;error\;of\;the\;mean
\\
t_{crit}\;=\;the\;critical\;value\;of\;the\;t\;statistic
$$

```{r}
###Calculate the mean of the hp
mean_hp <- mean(mtcars$hp)
print(paste0("Mean of horsepower: ",mean_hp))
###Calculate the sd of the hp
sd_hp <- sd(mtcars$hp)
print(paste0("Standard deviation of horsepoweer: ",sd_hp))

###Calculate the square root of the hp
n<- sqrt(length(mtcars$hp))
print(n)

###Confidence level of 0.95% e.g. two-tailed with 2.5%
t_value <- 1.96
###How to calculate the t-value properly
###Take the p-value: 0.05 and the degrees of freedom: 32-1
tval <- qt((1-0.95)/2, df=32-1)
print(paste0("t-critical value: ",tval))

###Standard error of the sample mean
se_sample_mean <- (sd(mtcars$hp)/sqrt(length(mtcars$hp)))
print(paste0("Standard Error of the Sample Mean",se_sample_mean))

sd(mtcars$hp)/sqrt(length(mtcars$hp))
ci_lower_bound <-mean(mtcars$hp)-(se_sample_mean*tval)

ci_upper_bound<- mean(mtcars$hp)+(se_sample_mean*tval)
print(ci_lower_bound)

print(ci_upper_bound)

# head(HumanResourcesDataAA5221_all_zscores[,c(1,11,2,12)]) %>%
#   kbl(row.names = T) %>%
#   kable_styling(row_label_position = "l", full_width = F) %>% 
#   footnote(general = "Head of data frame.") %>% 
#   add_header_above(c(" " = 1, "Satisfaction Level" = 2, "Last Evaluation" = 2))


```


## Utilizing the linear model formula in R {-}

## Here we can also calculate the $CI$ by utilizing the linear model funciton. {-}


```{r}
mean(mtcars$hp)+(1.96*12.1203)
# Calculate the mean and standard error
l.model <- lm(hp ~ 1, mtcars)
summary(l.model)

# Calculate the confidence interval
confint(l.model, level=0.95)

```

---

<center><h2>References</h2></center>

Hatcher, L. (2013). *Advanced statistics in research: Reading, understanding, and writing up data analysis results. Shadow Finch Media.*

https://www.statology.org/t-distribution-table/

https://www.scribbr.com/frequently-asked-questions/critical-value-of-t-in-r/#:~:text=You%20can%20use%20the%20qt,the%20significance%20level%20by%20two.
