# Example data to follow along with {-}


Detroit Data Set: A data frame with  14 Columns with 13 Observations

This is the data set called 'DETROIT' in the book 'Subset selection in regression' by Alan J. Miller published in the Chapman & Hall series of monographs on Statistics & Applied Probability, no. 40. The data are unusual in that a subset of three predictors can be found which gives a very much better fit to the data than the subsets found from the Efroymson stepwise algorithm, or from forward selection or backward elimination. The original data were given in appendix A of `Regression analysis and its application: A data-oriented approach’ by Gunst & Mason, Statistics textbooks and monographs no. 24, Marcel Dekker. It has caused problems because some copies of the Gunst & Mason book do not contain all of the data, and because Miller does not say which variables he used as predictors and which is the dependent variable. (HOM was the dependent variable, and the predictors were FTP … WE)

Source: http://lib.stat.cmu.edu/datasets/detroit

---

The necessary files to run statistical analyis on the Detroit data set presented in this book.

```{r echo=FALSE}
xfun::embed_dir('detroit/', text = 'Download Detroit data')
```

---

Student Performance Data Set: A data frame with 392 rows and 33 variables:

This data approach looks at student achievement in secondary education of two Portuguese schools. The data attributes include student grades, demographics, social and school related features and it was collected utilizing school reports and questionnaires. Two datasets are provided regarding the performance in two distinct subjects: Mathematics (mat) and Portuguese language (por). In [Cortez and Silva, 2008], the two datasets were modeled under binary/five-level classification and regression tasks. Important note: the target attribute G3 has a strong correlation with attributes G2 and G1. This occurs because G3 is the final year grade (issued at the 3rd period), While G1 and G2 correspond to the 1st and 2nd period grades. It is more difficult to predict G3 without G2 and G1, but such prediction is much more useful (see paper source for more details)

Source: http://archive.ics.uci.edu/ml/datasets/Student+Performance

---

The necessary files to run statistical analysis on the Student data set presented in this book.

```{r echo=FALSE}
xfun::embed_dir('student/', text = 'Download Student data')
```

---


Mtcars Data Set: A data frame with 32 rows and 12 variables:

The data was extracted from the 1974 Motor Trend US magazine, and comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973–74 models).

---

The necessary files to run statistical analysis on the mtcars data set presented in this book.

```{r echo=FALSE}
xfun::embed_dir('mtcars/', text = 'Download mtcars data')
```

---