# Partial Correlation


```{r echo=FALSE, include=TRUE}

library(mathjaxr)
library(latexpdf)
knitr::opts_chunk$set(echo = TRUE)

```

---

## Partial Correlation Assumptions {-}

Partial correlation allows us to control for variables that may be confounding variables in a data set. This allows us to see what the correlation between two variables would be if we control for a single variable. We can also control for more than one variable.

+ *First order partial coefficient* – is a correlation between two variables with just *one* additional variable partialed out of both.
+ *Higher order partial correlation* – is a correlation between two variables with *more than one control variable* partialed out by both.

## Partial Correlation Formula {-}

$$\Huge
r_{xy,z} = \frac{r_{xy} - r_{xz}  \cdot r_{yz}}{\sqrt{(1-r_{xz}^2) \cdot (1 - r_{yz}^2)}}$$



## Partial Correlation Overview {-}

When performing a single-order partial correlation we can simple do this by running a correlation on each variable in question. For our purposes we are utilizing the following variables horsepower (hp), mile pers gallon (mpg) and lastly cylinders (cy). If you would like to follow along we are utilizing the mtcars data set that is built into R.

+ Predictor (X) (Independent Variable) = hp (horsepower)
+ Criterition (Y) (Dependent Variable) = mpg (miles per gallon)
+ Control (Z) (Control Variable) = cyl (cylinders)

---

<center>
```{r,message=FALSE,echo=FALSE,out.width="35%",out.height="35%"}
library(DiagrammeR)
a_graph <-
  create_graph() %>%
  add_node(label = "X (hp)") %>%
  add_node(label = "Y (mpg)") %>%
  add_node(label = "Z (cyl)") %>%
  add_edge(from = 1, to = 2) %>% 
  add_edge(from = 3, to = 1) %>% 
  add_edge(from = 3, to = 2 )
render_graph(a_graph)
# mermaid("
# graph LR
#       X--Z;
#       Z--Y;
#       X--Y;
# ",width="300px",height="300px")
```
</center>

---


## Take a look at the structure of the mtcars data set {-}

```{r, echo=FALSE}
str(mtcars)
```

## Description of the variables in the mtcars data set

A data frame with 32 observations on 11 (numeric) variables.

+ Column 1	 =  mpg	- Miles/(US) gallon
+ Column 2	 =  cyl	- Number of cylinders
+ Column 3	 =  disp - Displacement (cu.in.)
+ Column 4	 =  hp - Gross horsepower
+ Column 5	 =  drat - Rear axle ratio
+ Column 6	 =  wt - Weight (1000 lbs)
+ Column 7	 =  qsec - 1/4 mile time
+ Column 8	 =  vs - Engine (0 = V-shaped, 1 = straight)
+ Column 9	 =  am - Transmission (0 = automatic, 1 = manual)
+ Column 10  =	gear - Number of forward gears
+ Column 11  =	carb - Number of carburetors

## Partial Correlation Calculation


### Correlation between Predictor and Criterion or X and Y

Here we find the correlation between our mpg and hp

```{r}
cor.test(mtcars$mpg,mtcars$hp)

```


### Correlation between Predictor and Control or X and Z

Here we find the correlation between our hp and cyl

```{r}
cor.test(mtcars$hp,mtcars$cyl)

```


### Correlation between Criterion and Control or Y and Z

Here we find the correlation between our mpg and cyl

```{r}
cor.test(mtcars$mpg,mtcars$cyl)

```


### Partial Correlation: Putting it altogether


```{r}
###Partial Correlation when controlling for another variable
###X is hp
###Y is mpg
###Z is cyl
xy <- cor(mtcars$hp,mtcars$mpg)
xz <- cor(mtcars$hp,mtcars$cyl)
yz <- cor(mtcars$mpg,mtcars$cyl)

print(xy)
print(xz)
print(yz)

```


### Calculate the numerator of the formula

$$\large
r_{xy} - r_{xz}  \cdot r_{yz}
$$

$$\large
-0.7761684 - (0.8324475  \cdot -0.852162)
$$

```{r}
###Top of the Formula
top_of_formula<- xy - (xz*yz)
print(top_of_formula)
```


### Calculate the denominator of the formula


$$\large
\sqrt{(1-r_{xz}^2) \cdot (1 - r_{yz}^2)}
$$


$$\large
\sqrt{(1-0.8324475^2) \cdot (1 - (-0.852162)^2)}
$$


```{r}
###Bottom of formula
bottom_of_formula<- sqrt((1-(xz^2))*(1-(yz^2)))
print(bottom_of_formula)
```

```{r}
final_answer= top_of_formula/bottom_of_formula
print(final_answer)
print(round(final_answer,2))
```
<b>Note: Round up at the very end of the calculation.</b>

$$\large
r = \frac{-0.06678832}{0.2899505}
$$
$$\large
r = -0.230439
$$

## Partial Correlation utilizing the ppcor package


**First order partial coefficient** – is a correlation between two variables with just one additional variable partialed out of both.

Here we can utilize the ppcor package as an easy button method to calculate the partial correlation. You will need to install the ppcor package via CRAN: [ppcor package](https://cran.r-project.org/web/packages/ppcor/index.html)

```{r}
###Partial Correlation when controlling for another variable
##Load the ppcor library
library(ppcor)


###ppcor function controlling for cylinders in the mtcars dataset
ppcor::pcor.test(mtcars$mpg, mtcars$hp, mtcars[, c("cyl")])
```

## Higher order partial correlation – is a correlation between two variables with more than one control variable partialed out by both.
 


```{r}
###ppcor function controlling for cylinders,displacement in the mtcars dataset
ppcor::pcor.test(mtcars$mpg, mtcars$hp, mtcars[, c("cyl","disp")])
```
